Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow

### 2024-05-13 manuscript/Manuscript.qmd
Update Manuscript.qmd
Mariannes endringsforslag

---
title: Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
  - name: Marianne Evju
    email: marianne.evju@nina.no
    orchid: https://orcid.org/0000-0001-7338-5376
    affiliations:
        - id: nina2
          name: Norwegian Institute for Nature Research
          department: NINA Oslo
          address: Sognsveien 68
          city: Oslo
          postal-code: NO-0855
abstract: |
  Ecosystem Condition Accounts (ECA) should reflect the integrity or quality of all nature inside the scope of the account, and therefore rely on spatially representative indicators for condition. All ECAs are subject to data constraints in some way. Therefore, being able to make use of spatially biased data sets would be very valuable. For national ECAs, modelling aproaches can in some cases be used to control for sampling biases. For local ECAs however, like at the scale of individual municipalities of developemetn projects, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area. In this study we develop three ecosystem condition indicators from the same spatially biased data set based on nature type mapping of Norwegian mires. The indicators are Alien species, Trenching and Anthropogenic Disturbance to Soil and Vegetation. We test our approach in three municipalities in south Norway. A spatial variable was created with four steps of increasing of infrastructure prevalence that we refer to as Homogeneous Impact Areas (HIAs). To facilitate religable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA seperately. Then we use area-weighted resampling and produce indicator propabiliy distributions for each municipality. With this Bayesian updating and resampling approach, small sample sizes can be compensated by correspondingly large uncertainty ranges, as long as the full dataset is large enough to estimate the true population standard deviation. This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey so that the data can be used for indicator in ECAs. The workflow can be used at any scale, including national scale. Because indicator values are calculatind for unique spatial strata, local governments or others can target their data aquisition towards strata with low sample sizes, and thus achieve higher cost effectiveness and ultimately better spatial indicator coverage.
keywords: 
  - alien species
  - disturbance
  - ecosystem accounting
  - ecosystem condition
  - horizontal aggregation
  - indicator
  - mire
  - peatlands
  - SEEA EA
  - wetlands
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the EAs [@unitednations2021 §5.54].
Ecosystem assets are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies systematically along gradients of anthropogenic pressures and hence ecosystem condition.
This recommendation for the spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to to reliably estimate an indicator value for each EA.
This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each EA will have ebough data to estaimte its unique indicator value.
This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer*[@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
There are at least three ways to achieve this same complete areal coverage from indicator values that originally are spatially incomplete:

@@

The need for an unbiased estimation of indicator values is unquestionable, but nonetheless, this requirement puts a large limitation on what types of data one can use.
Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments are more often than not a pragmatic and opportunistic one, which is unlikely to reflect the full scope of the ecosystem condition characteristics.
Note that the same is true for thematic biases.
This is for example reflected in the scarcity of data commonly included on insects or soil biota, even though most will agree they represent key ecosystem characteristics.
Also having data from only one or a subset of nature types inside what is defined as the ecosystem in the assessment, is a typical thematic bias in ecosystem accounting.
However, in this paper we chose to focus on spatial bias.

Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
One way to achieve this is modelling (option *a* in the list above).
Statistical models can describe the general associations between the sampled data and the context where it was sampled (e.g. a set of environmental variables), and use these relationships to predict and project indicator values to areas that where not originally sampled.
Depending on the data that goes into these models, they can make very precice and good indicators.
This is especially true when the EAAs are large (e.g. regions or nations).
But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from the spatial extrapolation of such models becomes unacceptable.

@@

In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones [@asbjørnmoen1998].
In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
We included data from 2018 to 2023.
In this paper we assume the survey is representative for the entire mire ecosystem in Norway (i.e. that there is no thematic bias).
Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed, we do not think this is so much the case for our variables (see below).
However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will be biased since that variable for only recorded in fens.
The other variables are recorded in all delineated mires.

From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
Variable 1, 7FA, represents the abundance of alien species, mainly plants.
Variables 2-5 (7SE, PRSL, 7TK and PRTK) describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they where combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
These variables were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
This was done by summing the four variables after they had been converted to percentages.
This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
Variable 6, 7GR-GI, is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
This is not a favourable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.

To turn variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition [@jakobsson2020].
The reference values makes up a numerical representation of the reference condition.
We define the reference condition as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today [@jakobsson2021].

The variable 7TK was rescaled into the indicator named *Alien species*, 7GR-GI was rescaled into the indicator *Trenching*, and ADSV was rescaled and kept the name *ADSV* (Figure S1).
For *Alien species* and *ADSV*, the lower and upper reference values were defined as 100% and 0%, respectively.
The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
For *Trenching* we used the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighbouring (ecologically speaking) mapping unit.
See @fig-workflowExample and S1 for schematic workflows for the indicator development.
*Alien species* was attributed to Ecosystem Condition Typology (ECT) class B1 - Compositional state characteristics and *ADSV,* and *Trenching* was attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].

@@

Because the nature type survey data are spatially biased, we cannot assume that they are area-representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel [@erikstad2023].
We then categorised this continuous variable into four classes (0–3) using ranges chosen to produce a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each strata across Norway (Table S1).
We then aggregated the data to 1×1 km pixels to ease computations, and vectorized it.
The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.

We used the entire national data set for assessing the relationship between the HIA levels and the indicator values.
We then sub-setted the national data set and chose three municipalities in south Norway to test out the indicators.
The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, S3, S4).
We then took the nature type polygons with the indicator values and intersected with the HIA map and a map of municipal outlines in Norway.
The relationship between the HIA classes and indicator values was examined visually.
For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set (@fig-workflowExample C; Appendix C, line 138 <!--# update line number manually when re-rendering appendix C -->).
We refer to this step as the first of two *aggregation* steps (Figure S1).
The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately.
This step is referred to as *spreading* the data (Figure S1).
For each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
The number of m^2^ dictated the number of samples for each polygon, thus ensuring that large polygons ended up counting more towards the indicator value in the given municipality.
We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for each municipality (the second aggregation step; Figure S1).
When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.

@@

```{r fig-workflowExample}
#| eval: true
#| include: true
#| out.width: '100%'
#| fig.cap:
#| - "Simplified schematic showing the workflow (Panes A-E) for spatial generalisation of a spatially biased ecosystem condition indicator. Pane A shows a spatially explicit, patchy and spatially biased indicator. The outline is the ecosystem accounting area (EEA). Pane B shows the location of four homogeneous impact areas (HIAs) inside the EEA. The indicator values in Pane A are used, in combination with the HIA map in Pane B, to update a uniform prior and produce a posterior probability distribution for the mean area woeghted indicator value for each HIA (Pane C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)). The bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set. The width of the posterior distribution responds both to variation in the data and to the sample size, giving  a realistic measure for the uncertainty around the indicator even with a singe observation. The distributions in Pane C are assumed normally distributed. Note that because in this example there were no indicator data for HIA-3, indicator values are only aggregated for HIAs 0–2. In Pane D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately. The colors indicator different indicator values (highlighted in the inset). For each ecosystem occurence (i.e. ecosystem assets) in Pane D, we draw one random value for each square meter of area from a normal distribution with the mean and SD that is associated with that polygon. In Pane E we have randomly sampled 1000 values from the entire vector of samples in the previous step, and we get an area weighted probability distribution for the mean indicator value for the EEA."

@@

```{r fig-validation}
#| eval: true
#| include: true
#| out.width: '90%'
#| fig.cap:
#| - "Proportion of localities with different indicator values for three 
#| ecosystem condition indicators along an ordinal gradient of increasing
#| infrastructure densities (Homogeneous Impact Areas 0–3). 
#| The data is from a national vegetation type survey in Norway."

@@

```{r fig-spread-example}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "An indicator for mire trenching shown for Nord-Aurdal municipality. 
#| Individual mire polygons are colored by the mean indicator values for the
#| homogenous impact area where it lies. Colors are chosen to best reflect 
#| categorical differences and exagerates the absolute difference between 
#| areas. The inset it just a visual aid. Coordinate reference system is 
#| EPSG 25832. Axis are in meters."

@@

```{r fig-EEA-ridgePlot}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Distributions for three ecosystem condition indicators in the Norwegian 
#| municiaplities. The color gradient reflects the value of the x-axis. 
#| The dotted vertical line represent the threshold for what is considered 
#| reduced ecosystem condtion (< 0.6). Vertical lines under the density curves
#| are 2.5%, 50% (the median) and 97.5% percentiles."

@@

In this paper we have demonstrated a generalisable GIS-based workflow for making use of spatially biased field data to produce representative ecosystem condition indicators at a local municipal scale (@fig-workflowExample).
This approach is useful especially for local assessments were we do not want to use data from outside the EAA inform or influence our estimates for condition indicators inside the EAA.
The method relies on stratifying the field data by Homogeneous Impact Areas (HIAs).
Our HIA was an ordinal gradient from low to high prevalence of infrastructure, and we found a relationship of increasing infrastructure and decreasing indicator values ranging from weak, in the case of the *Alien species* indicator, to strong, for the indicator for *Trenching* in mires (the *ADSV* indicator falling somewhere in the middle) (@fig-validation) .
This implies that the stratification was warranted in the latter case (*Trenching*), i.e that field data from urban or sub-urban mires could not be said to represent the status of mires in more remote wilderness areas.
In the case for *Alien species*, stratifying the variable by the HIAs can be predicted to have had a less directional effect on average, but we did find a difference between HIA-1 and 2 in Nord-Aurdal municipality, indicating that the impact from alien species differs between these two strata of varying infrastructure (@fig-forest).
The drawback that we see from using this stratification technique when the HIA variable is not a good predictor for variation in the indicator, is that we end up with less data points, and therefore higher uncertainties around our indicator estimates, than we otherwise would.
However, we do not see this as a major problem.
Also, since our HIA areas are somewhat cohesive, forming a clumped or aggregated pattern, stratifying by it will on average lead to relatively more local data being used for each local indicator estimate, with less geographic extrapolation.
Therefor, we see little risk in using the stratification process also on indicators where the relationship between the indicators and the HIA are weak, although we stress that not accounting for other, potentially more relevant predictors of indicator values, would leave the indicator as spatially biased as the original variable.

@@

We have also demonstrated how Bayesian updating and re-sampling can be used instead of traditional arithmetic aggregation of indicators, from EAs to EAAs, to produce probability density distributions for indicator values.
A major benefit of this approach in our example was that it allowed us to present reliable uncertainty ranges for our indicator estimates even with very low sampling sizes, thus making it possible to estimate indicator values for a much greater areas than if we were forced to set a minimum sample size threshold.
An indicator estimate is of little value unless it is accompanied by a measure of uncertainty.
The variance is commonly used, usually calculated from the variation between spatially distant sampling points.
However, the variance is not reliable for low sample sizes.
For example, even if a numerical population has a large variance, it is still possible to sample, by chance alone, two or three numbers that are very close to each other, and this sample would then display a low variance which is not reflective of the true variance.
Knowing this, an often used approach is to set a sample size limit, and refrain from estimating indicator values unless there are more than a given number of data points.
In our test municipality Nordre Follo, we had only 1-3 data points for each HIA strata, yet we were able to use these data regardless because the Bayesian updating produces a reliable, and relatively broad, uncertainty range for the estimate.
This is an advantage, especially for local ecosystem condition assessments where data scarcity is likely to be more of an issue than in for example national assessments.

@@

Alien species risk is not considered a threat to any of the Norwegian mire types , and our results confirm this, at least for the present (@fig-EEA-ridgePlot).
Motorised vehicle traffic and other forms of human transportation are also threats listed for some mire types in the red list assessment, and we see some reduction in the ADSV indicator, most so in Gran and in areas closer to infrastructure (@fig-forest).
Still, the indicator values are consistently above the threshold value (@fig-EEA-ridgePlot).

@@

**Anders L. Kolstad**: Conceptualization, Methodology, Software, Validation, Formal analysis, Data Curation, Writing - Original Draft, Visualization, Project administration, Funding acquisition.
**Matthew Grainger**: ...
**Marianne Evju**: ...

>>>>

---
title: Making use of spatially biased variables in ecosystem condition accounting – a GIS based workflow
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Matthew Grainger
    email: matthew.grainger@nina.no
    orchid: https://orcid.org/0000-0001-8426-6495
    affiliations:
        - ref: nina
  - name: Marianne Evju
    email: marianne.evju@nina.no
    orchid: https://orcid.org/0000-0001-7338-5376
    affiliations:
        - id: nina2
          name: Norwegian Institute for Nature Research
          department: NINA Oslo
          address: Sognsveien 68
          city: Oslo
          postal-code: NO-0855
abstract: |
  Ecosystem Condition Accounts (ECA) should reflect the integrity or quality of all nature inside the scope of the account, and therefore rely on spatially representative indicators for condition. All ECAs are subject to data constraints in some way. Therefore, being able to make use of spatially biased data sets would be very valuable. For national ECAs, modelling aproaches can in some cases be used to control for sampling biases. For local ECAs however, like at the scale of individual municipalities of developement projects, this is often not an option as it involved spatial extrapolation using data from outside the ecosystem accounting area. In this study we develop three ecosystem condition indicators from the same spatially biased data set based on nature type mapping of Norwegian mires. The indicators are Alien species, Trenching and Anthropogenic Disturbance to Soil and Vegetation. We test our approach in three municipalities in south Norway. A spatial variable was created with four steps of increasing of infrastructure prevalence that we refer to as Homogeneous Impact Areas (HIAs). To facilitate reliable estimation of indicator uncertainties, even in cases with very low sample sizes, we use a Bayesian updating method and produce probability distributions for the area-weighted mean indicator values in each HIA seperately. Then we use area-weighted resampling and produce indicator propabiliy distributions for each municipality. With this Bayesian updating and resampling approach, small sample sizes can be compensated by correspondingly large uncertainty ranges, as long as the full dataset is large enough to estimate the true population standard deviation. This paper demonstrates the use of a GIS-based workflow to control for some of the most problematic biases in an oppurtunistic field survey so that the data can be used for indicators in ECAs. The workflow can be used at any scale, including national scale. Because indicator values are calculatind for unique spatial strata, local governments or others can target their data aquisition towards strata with low sample sizes, and thus achieve higher cost effectiveness and ultimately better spatial indicator coverage.
keywords: 
  - alien species
  - disturbance
  - ecosystem accounting
  - ecosystem condition
  - horizontal aggregation
  - indicator
  - mire
  - peatlands
  - SEEA EA
  - wetlands
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
  warning: false
  message: false
---

@@

The SEEA EA framework recommends that ecosystem condition indicators give an unbiased representation of the condition for a given area [table 1 in @czucz_selection_2021; see also @unitednations2021 §2.87], and that condition indicators are recorded uniquely for each ecosystem asset (EA) so that the condition in the ecosystem accounting area (EAA) can be inferred from an area weighted mean based on the relative extents of the ecosystem assets [@unitednations2021 §5.54].
Ecosystem assets (EAs) are defined as "ecological entities" (meaning areas) about which information is sought and about which statistics are compiled [@unitednations2021].
This recommendation for representativity in the indicators means that spatially biased data are ill suited, especially if sampling intensity varies systematically along gradients of anthropogenic pressures and hence ecosystem condition.
The recommendation for spatial resolution also means that when a condition metric is sampled coarsely, for example in an opportunistic and spatially biased field campaign, then it is very likely that you will not have a enough data points to reliably estimate an indicator value for each EA.
This problem does not exist for complete wall-to-wall data, like remotely sensed imagery, where each EA will have ebough data to estaimte its unique indicator value.
This challenge if spatial generalisation is not unique to ecosystem condition accounts, and it is much discussed in the ecosystem service accounting community of practice via the related term *value transfer* [@unitednations2022; @ncaves; @grammatikopoulou2023; @barton2023].
There are at least three ways to achieve this same complete areal coverage from indicator values that originally are spatially incomplete:

@@

The need for an unbiased estimation of indicator values is unquestionable, but nonetheless, this requirement puts a large limitation on what types of data one can use in ecosystem condition accounts.
Ecosystem condition assessments are generally limited by data availability, and the choice of variables and indicators to include in assessments is more often than not a pragmatic and opportunistic one, which is unlikely to reflect the full scope of the ecosystem condition characteristics.
Note that the same is true for thematic biases.
This is for example reflected in the scarcity of data commonly included on insects or soil biota, even though most will agree they represent key ecosystem characteristics.
Also having data from only one or a subset of nature types inside what is defined as the ecosystem in the assessment, is a typical thematic bias in ecosystem accounting.
However, in this paper we chose to focus on spatial bias.

Being able to make use of spatially biased data would greatly alleviate data shortage problems in ecosystem condition accounts.
One way to achieve this is modelling (option *a* in the list above).
Statistical models can describe the general associations between the sampled data and the context where it was sampled (e.g. a set of environmental variables), and use these relationships to predict and project indicator values to areas that where not originally sampled.
Depending on the data that goes into these models, they can make very precice and good indicators (REF).
This is especially true when the EAAs are large (e.g. regions or nations).
But when they are small, like the scale of a municipality, and when the indicator is more likely to be used as the evidence base in concrete physical land use planning, then the inherent level of uncertainty from the spatial extrapolation of such models becomes unacceptable.

@@

In the above, *southern* refers to boreonemoral and southboreal zones, and *northern* refers to mid-boreal, north boreal, and alpine zones [@asbjørnmoen1998].
In addition, the northern fens need to be even more calcareous than the southern fens in order to be surveyed.
We included data from 2018 to 2023.
In this paper we assume the survey is representative for the entire mire ecosystem in Norway (i.e. that there is no thematic bias).
Although it is possible that smaller or less calcareous mires will score systematically different than the ones that are surveyed, we do not think this is so much the case for our variables (see below).
However, we do assume that alien plants are slightly less common on bogs relative to fens, and therefore that this variable will be biased since that variable is only recorded in fens.
The other variables are recorded in all delineated mires.

From the survey data set we identified six relevant variables (variable 1-6 in @tbl-variables) which we attribute to three different ecosystem condition characteristics that describe the typical behavior of open mire ecosystems in the reference condition (Figure S1).
Variable 1, 7FA, represents the abundance of alien species, mainly plants.
Variables 2-5 (7SE, PRSL, 7TK and PRTK) describe very related aspects, and are attributed to the same ecosystem condition characteristic (vegetation intactness) and so they were combined into a single indicator called anthropogenic disturbance to soil and vegetation (ADSV; Figure S1).
These variables were were originally recorded along binned frequency ranges @norwegian_environmental_agency_protocol .
Because the data was strongly right skewed we used the lower limit for each frequency range to convert them into percentages.
This was done by summing the four variables after they had been converted to percentages.
This was not a perfect solution, especially since some localities only had one of the variables recorded, but we chose this, rather than for example using a *worst rule* principle, in order to better separate the localities in terms of their indicator values.
Variable 6, 7GR-GI, is different from the preceding variables in @tbl-variables in that it includes an estimation of future effects that the observed trenches are projected to have on mire vegetation, function or structure over time.
This is not a favourable trait in a metric used to evaluate the ecosystem condition as it is today, yet we include it here nonetheless because there is a general shortage of data on mire hydrology, which is a fundamental part of mire integrity.

To turn variables into indicators we scaled them using three normative reference values each: an upper (best possible condition), a lower (worst possible condition), and a threshold value that defines the breaking point between good and poor condition [@jakobsson2020].
The reference values make up a numerical representation of the reference condition.
We define the reference condition as one where ecosystems are subject to little or no human influence, with a climate as in the period 1961-1990 and a native species pool similar as today [@jakobsson2021].

The variable 7FA was rescaled into the indicator named *Alien species*, 7GR-GI was rescaled into the indicator *Trenching*, and ADSV was rescaled and kept the name *ADSV* (Figure S1).
For *Alien species* and *ADSV*, the lower and upper reference values were defined as 100% and 0%, respectively.
The threshold for *good ecosystem condition* was defined as 10%, which was then mapped to the value 0.6 on the rescaled indicators, thus creating a non-linear rescaling of the variable (Figure S2).
For *Trenching* we used the lower and upper reference values 1 and 5, respectively, and a threshold value of 2.5 (Figure S2).
A variable value of 1 indicates an intact mire, and a value of 5 indicates a mire transitioning away from a wetland.
A value of 2 indicates observable change within the range expected for the same mapping unit, and a value of 3 indicates a mire transitioning into a neighbouring (ecologically speaking) mapping unit.
See @fig-workflowExample and S1 for schematic workflows for the indicator development.
*Alien species* was attributed to Ecosystem Condition Typology (ECT) class B1 - Compositional state characteristics and *ADSV,* and *Trenching* were attributed to ECT class A1 - Physical state characteristics [@czucz_common_2021].

@@

Because the nature type survey data are spatially biased, we cannot assume that they are area-representative.
In an attempt to overcome this issue we divided the area of Norway into four non-overlapping Homogeneous Impact Areas (HIAs) based on an infrastructure index for the year 2022 (@tbl-variables).
This index is a continuous variable that represents the frequency of different infrastructure types inside 500 m radius circles around each 100×100 m pixel [@erikstad2023].
We then categorised this continuous variable into four classes (0–3) using ranges chosen to produce a sensible and relatively linear area classification when visualised along an urban-wilderness gradient, whilst at the same time delineating sufficient areas for each strata across Norway (Table S1).
We then aggregated the data to 1×1 km pixels to ease computations, and vectorised it.
The name HIAs is not a perfect representation of the information found in the infrastructure index, but we introduce this name here as a general term which is aligned with the concept of Homogeneous Ecosystem Areas, *sensu* @vallecillo_eu-wide_2022.

We used the entire national data set for assessing the relationship between the HIA levels and the indicator values.
We then sub-setted the national data set and chose three municipalities in south Norway to test out the indicators.
The municipalities differ in several aspects, such as the amount of mire area, the total area surveyed, and the prevalence of infrastructure (@tbl-munis, @fig-studyLocation2, S3, S4).
The nature type polygons with the indicator values were intersected with the HIA map and a map of municipal outlines in Norway.
The relationship between the HIA classes and indicator values was examined visually.
For each HIA and municipality combination we then created a probability distribution for the area weighted mean of the indicator values using Bayesian updating of a uniform prior between 0 and 1, informed by the standard deviation (SD) of the indicator values in the national data set (@fig-workflowExample C; Appendix C, line 138 <!--# update line number manually when re-rendering appendix C -->).
We refer to this step as the first of two *aggregation* steps (Figure S1).
The resulting distributions are assumed to be normally distributed, and we therefore simply carried the mean and SD from the posterior distributions over to individual polygons in the ecosystem delineation map, for each HIA class separately.
This step is referred to as *spreading* the data (Figure S1).
For each polygon we then sampled random numbers from a normal distribution with this same mean and SD.
The number of m^2^ dictated the number of samples for each polygon, thus ensuring that large polygons ended up counting more towards the indicator value in the given municipality.
We then drew 1000 random values from this vector of possible indicator values and created a probability distribution for the indicators for each municipality (the second aggregation step; Figure S1).
When there were no indicator values for a given HIA class, we ignored that class also in the municipal estimate.

@@

```{r fig-workflowExample}
#| eval: true
#| include: true
#| out.width: '100%'
#| fig.cap:
#| - "Simplified schematic showing the workflow (Panes A-E) for spatial generalisation of a spatially biased ecosystem condition indicator. Pane A shows a spatially explicit, patchy and spatially biased indicator. The outline is the ecosystem accounting area (EEA). Pane B shows the location of four homogeneous impact areas (HIAs) inside the EEA. The indicator values in Pane A are used, in combination with the HIA map in Pane B, to update a uniform prior and produce a posterior probability distribution for the mean area weighted indicator value for each HIA (Pane C; here simplified to only show the mean (circles) and 95% credible intervals (colored bands)). The Bayesian updating is informed by the standard deviation for the indicator in a much bigger national data set. The width of the posterior distribution responds both to variation in the data and to the sample size, giving  a realistic measure for the uncertainty around the indicator even with a single observation. The distributions in Pane C are assumed normally distributed. Note that because in this example there were no indicator data for HIA-3, indicator values are only aggregated for HIAs 0–2. In Pane D the mean and SD from the posterior distributions are transferred to individual polygons in an ecosystem delineation map, for each HIA seperately. The colors indicate different indicator values (highlighted in the inset). For each ecosystem occurence (i.e. ecosystem assets) in Pane D, we draw one random value for each square meter of area from a normal distribution with the mean and SD that is associated with that polygon. In Pane E we have randomly sampled 1000 values from the entire vector of samples in the previous step, and we get an area weighted probability distribution for the mean indicator value for the EEA."

@@

```{r fig-validation}
#| eval: true
#| include: true
#| out.width: '90%'
#| fig.cap:
#| - "Proportion of localities with different indicator values for three 
#| ecosystem condition indicators along an ordinal gradient of increasing
#| infrastructure densities (Homogeneous Impact Areas 0–3). 
#| The data is from a national nature type survey in Norway."

@@

```{r fig-spread-example}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "An indicator for mire trenching shown for Nord-Aurdal municipality. 
#| Individual mire polygons are colored by the mean indicator values for the
#| homogenous impact area where it lies. Colors are chosen to best reflect 
#| categorical differences and exaggerates the absolute difference between 
#| areas. The inset it just a visual aid. Coordinate reference system is 
#| EPSG 25832. Axis are in meters."

@@

```{r fig-EEA-ridgePlot}
#| eval: true
#| include: true
#| out.width: '80%'
#| fig.cap:
#| - "Distributions for three ecosystem condition indicators in the Norwegian 
#| municipalities. The color gradient reflects the value of the x-axis. 
#| The dotted vertical line represents the threshold for what is considered 
#| reduced ecosystem condtion (< 0.6). Vertical lines under the density curves
#| are 2.5%, 50% (the median) and 97.5% percentiles."

@@

In this paper we have demonstrated a generalisable GIS-based workflow for making use of spatially biased field data to produce representative ecosystem condition indicators at a local municipal scale (@fig-workflowExample).
This approach is useful especially for local assessments were we do not want to use data from outside the EAA to inform or influence our estimates for condition indicators inside the EAA.
The method relies on stratifying the field data by Homogeneous Impact Areas (HIAs).
Our HIA was an ordinal gradient from low to high prevalence of infrastructure, and we found a relationship of increasing infrastructure and decreasing indicator values ranging from weak, in the case of the *Alien species* indicator, to strong, for the indicator for *Trenching* in mires (the *ADSV* indicator falling somewhere in the middle) (@fig-validation) .
This implies that the stratification was warranted in the latter case (*Trenching*), i.e that field data from urban or sub-urban mires could not be said to represent the status of mires in more remote wilderness areas.
In the case for *Alien species*, stratifying the variable by the HIAs can be predicted to have had a less directional effect on average, but we did find a difference between HIA-1 and 2 in Nord-Aurdal municipality, indicating that the impact from alien species differs between these two strata of varying infrastructure (@fig-forest).
The drawback that we see from using this stratification technique when the HIA variable is not a good predictor for variation in the indicator, is that we end up with less data points, and therefore higher uncertainties around our indicator estimates, than we otherwise would.
However, we do not see this as a major problem.
Also, since our HIA areas are somewhat cohesive, forming a clumped or aggregated pattern, stratifying by it will on average lead to relatively more local data being used for each local indicator estimate, with less geographic extrapolation.
Therefore, we see little risk in using the stratification process also on indicators where the relationship between the indicators and the HIA are weak, although we stress that not accounting for other, potentially more relevant predictors of indicator values, would leave the indicator as spatially biased as the original variable.

@@

We have also demonstrated how Bayesian updating and re-sampling can be used instead of traditional arithmetic aggregation of indicators, from EAs to EAAs, to produce probability density distributions for indicator values.
A major benefit of this approach in our example was that it allowed us to present reliable uncertainty ranges for our indicator estimates even with very low sampling sizes, thus making it possible to estimate indicator values for a much greater area than if we were forced to set a minimum sample size threshold.
An indicator estimate is of little value unless it is accompanied by a measure of uncertainty.
The variance is commonly used, usually calculated from the variation between spatially distant sampling points.
However, the variance is not reliable for low sample sizes.
For example, even if a numerical population has a large variance, it is still possible to sample, by chance alone, two or three numbers that are very close to each other, and this sample would then display a low variance which is not reflective of the true variance.
Knowing this, an often used approach is to set a sample size limit, and refrain from estimating indicator values unless there are more than a given number of data points.
In our test municipality Nordre Follo, we had only 1-3 data points for each HIA strata, yet we were able to use these data regardless because the Bayesian updating produces a reliable, and relatively broad, uncertainty range for the estimate.
This is an advantage, especially for local ecosystem condition assessments where data scarcity is likely to be more of an issue than in for example national assessments.

@@

Alien species risk is not considered a threat to any of the Norwegian mire types, and our results confirm this, at least for the present (@fig-EEA-ridgePlot).
Motorised vehicle traffic and other forms of human transportation are also threats listed for some mire types in the red list assessment, and we see some reduction in the ADSV indicator, most so in Gran and in areas closer to infrastructure (@fig-forest).
Still, the indicator values are consistently above the threshold value for good ecological condition (@fig-EEA-ridgePlot).

@@

**Anders L. Kolstad**: Conceptualization, Methodology, Software, Validation, Formal analysis, Data Curation, Writing - Original Draft, Visualization, Project administration, Funding acquisition.
**Matthew Grainger**: ...
**Marianne Evju**: Conceptualization, Writing - Review & Editing
